---
title: "post_hoc_analysis"
output: html_document
date: "2026-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
custom_colors <- c('#91bfdb', '#ffffbf','#fc8d59', "#a8ddb5") 
```

# Post-hoc analyses for WINDEX output 

The WINDEX output files themselves are hard to navigate, especially at the whole-genome level when they are rather large. Here are some example analyses that can be done with WINDEX output to make your research life easier! 

## window-level analyses
### bring in testing data 

```{r}
pattern_range <- "(8[1-9]|9[0-9]|100)\\.window_classified$"
file_list <- list.files(pattern = pattern_range, full.names = T)

window_results_list <- list()

for (i in 1:length(file_list)) {
  window_results_list[[file_list[i]]] <- read.csv(file_list[i], sep = '\t')
}

all_window_results <- bind_rows(window_results_list, .id = "source_file") %>% mutate(sim = str_extract(source_file, "[0-9]+"))
```

### generate confusion matrix at window-level 

Based on 20 testing simulations with 25 windows per simulation, we expect to see 20 sweep windows, 40 linked windows, and 440 neutral windows. 

```{r}
# make "true label" column for confusion matrix 
all_window_results_labels <- all_window_results %>% # first, make class variable so that WINDEX observations can be interpreted as classes
  mutate(pred_name = ifelse(pred == 1, "neutral", 
                 ifelse((pred == 2 | pred == 4), "linked", 
                 "sweep"))) %>% 
  mutate(label = ifelse(win == 480001, "sweep", # next, add in true labels based on training scheme (see Methods)
                 ifelse((win == 440001 | win == 520001), "linked", 
                 "neutral")))
```

```{r}
# confusion matrix
conf_matrix_table <- table(Actual = all_window_results_labels$label, Predicted = all_window_results_labels$pred_name)
print(conf_matrix_table)
```

### plot window results over all testing simulations 

```{r}
# summarize the data
all_window_results_plot <- all_window_results_labels %>% 
  group_by(win, pred_name) %>% 
  summarise(count = n())
```

```{r}
# plot
ggplot(all_window_results_plot, aes(x = win, y = (count/20), fill = factor(pred_name, levels = c("neutral", "linked", "sweep")))) + 
  geom_col() + 
  theme_light() + 
  labs(x = "window", y = "proportion of simulations", fill = "WINDEX prediction") + 
  scale_fill_manual(values = custom_colors) + 
  theme(axis.text.x = element_text(angle=90)) 
```

## site-level analyses
### bring in testing data 

```{r}
pattern_range <- "(8[1-9]|9[0-9]|100)\\.site_classified$"
file_list <- list.files(pattern = pattern_range, full.names = T)

site_results_list <- list()

for (i in 1:length(file_list)) {
  site_results_list[[file_list[i]]] <- read.csv(file_list[i], sep = '\t')
}

all_site_results <- bind_rows(site_results_list, .id = "source_file") %>% mutate(sim = str_extract(source_file, "[0-9]+"))
```

### analyze the number of sweep loci identified correctly in positively IDed sweep windows 

Because of the WINDEX state structure, the only circumstance where a sweep site will be identified in the data is if it is located inside a sweep window. Therefore, when analyzing the "success" of WINDEX at the site-level, we look to the simulations where WINDEX has sweep classifications in our true sweep window only. 

```{r}
# get simulations where sweep window was correctly identified 
sims_correct_sweep_window <- all_window_results %>% filter(pred == 3 & win == 480001) 

# filter for these simulations only 
all_site_results_filtered <- all_site_results %>% filter(sim %in% sims_correct_sweep_window$sim)
```

Next, let's add in our `pred_name` variable similar to the window-level case, but since we have a lot of sites to visualize even in our 40kb sweep window, let's only look at what is happening +/- 1kb from the true sweep site. We will do this by filtering for these sites by position. Since each simulation/genome file doesn't have the exact same sites or number of sites in them, we can bin sites into bins of 20 for visualization purposes. 

```{r}
bin_numbers <- seq(499000, 501000, 20) # set bins of 20 over the +/- 1kb region

# make pred_name variable and get counts in each bin
all_site_results_filtered_cleaned <- all_site_results_filtered %>% 
  mutate(bins = cut(as.numeric(site), breaks = bin_numbers)) %>% 
  mutate(pred_name = ifelse((site_pred == 1 | site_pred == 3), "linked", "sweep")) %>% 
  mutate(label = ifelse(site == 500001, "sweep", "linked")) %>% 
  group_by(bins, pred_name) %>% 
  summarise(count = n())

# get prediction counts over all bins
pred_counts <- all_site_results_filtered %>% mutate(bins = cut(as.numeric(site), breaks = bin_numbers)) %>% group_by(bins) %>% summarise(pred_count = n())

# merge site results and total prediction counts
plot_data <- left_join(all_site_results_filtered_cleaned, pred_counts, by = c("bins")) %>% mutate(prop = count / pred_count) %>% filter(!is.na(bins))

ggplot(plot_data, aes(x = bins, y = prop, fill = as.factor(pred_name))) + geom_col() + theme_minimal() + labs(x = "position", y = "proportion of class", fill = "class", title = "WINDEX site classifications in sweep simulations, +/- 1kb") + theme(axis.text.x = element_blank()) + scale_fill_manual(values = c('#ffffbf','#fc8d59'))
```

